### 2009-04-20 paper.tex
further fixed up null cycle section

Null cycles in SCFGs constitute a ``missing data'' problem, wherein
the unobserved values of variables contribute to probabilities of
observed events. Such missing data problems are common in
probabilistic models of sequences and can
frequently be dealt with by exactly marginalizing over the
probabilities of the unobserved variables.  For example, eliminating
missing ``empty'' columns from an independent-sites model of sequence,
wherein alignment columns are independent, is straightforward
\cite{RivasEddy2008} and missing emissions in a HMM can be dealt with by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007} or the ``wing retraction''
techniques used in HMMER \cite{hmmer}.
Algorithms for exact elimination of HMM null cycles by
matrix inversion are well-known \cite{Holmes2003} and are implemented
by programs such as PhyloComposer \cite{Holmes2007} and
HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
with an equivalent HMM that contains no null cycles but generates the
same probability distribution over sequences.
However, these solutions do not easily generalize to SCFGs, due to the
possible presence of null bifurcations.

Null cycles can be removed from SCFGs by (i) separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children and then (ii) encapsulating all paths that do
not emit any terminal symbols (and may therefore cause cyclic
dependencies in the DP matrix) as effective direct transitions, with
probabilities obtained by geometric series summation of the grammar's
transition matrix.

Both (i) empty bifurcations and (ii) paths without terminal symbols
generically appear in the ensemble model.
An empty bifurcation occurs when a child branch of a bifurcation state
transitions to the $\Send$ state without emitting any symbols and can
be removed from the model by creating an effective direct transition
encapsulating the empty bifurcation.  For
example, we can create an effective direct transition $N_1 \rightarrow
N_3$ between null states $N_1$ and $N_3$ in place of the empty
bifurcation $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
(End \,\, N_3)$, where $B$ is a bifurcation state with children $(N_2
\,\, N_3)$. Bifurcation states are the most computationally-costly
part of our models, so it is important to eliminate as many as
possible without reducing model expressiveness.

Similarly, our model construction procedure generically produces many
``windback'' null transitions to non-emitting placeholder
states. If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph by $\Snull$ states $N_1$ and $N_2$, then the transition $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced with a transition $E_1 \rightarrow
E_2$ with an identical probability.

@@

>>>>

The ensemble model contains empty bifurcations {\em and} empty paths.
An {\bf empty bifurcation} occurs when a child branch of a bifurcation state
transitions to the $\Send$ state without emitting any symbols and can
be removed from the model by creating an effective direct transition
encapsulating the empty bifurcation.  For
example, we can create an effective direct transition $N_1 \rightarrow
N_3$ between null states $N_1$ and $N_3$ in place of the empty
bifurcation $N_1 \rightarrow B \rightarrow (N_2 \,\, N_3) \rightarrow
(End \,\, N_3)$, where $B$ is a bifurcation state with children $(N_2
\,\, N_3)$. Bifurcation states are the most computationally-costly
part of our models, so it is important to eliminate as many as
possible without reducing model expressiveness.

Similarly, our model construction procedure generically produces many
{\bf empty paths} which do not emit any terminal symbols.
Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph by $\Snull$ states $N_1$ and $N_2$, then the transition $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced with a transition $E_1 \rightarrow
E_2$ with an identical probability.

@@

We refer to empty paths and empty bifurcations as {\bf null cycles}.
They occur (in simpler form) in Hidden Markov Models (which are special cases of SCFGs)
and independent-sites models (which can be viewed as special cases of HMMs).
In all such models, null cycles represent events which are allowed by the model, but could never be observed.
They can be accounted for by exact marginalization.
For example, eliminating missing ``empty'' columns from an independent-sites model of sequence involves a straightforward application of a correction factor
to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all null paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}; the ``wing retraction'' algorithms used in HMMER can be viewed as a special case of matrix inversion by iterated elimination \cite{hmmer}.

Algorithms for exact elimination of HMM null cycles by matrix inversion are implemented
by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
with an equivalent HMM that contains no null cycles but generates the
same probability distribution over sequences.
However, these solutions do not easily generalize to SCFGs, due to the
possible presence of null bifurcations.

Here, we show how null cycles can be removed systematically from SCFGs by two steps:
\begin{description}
\item[(i)] separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children;
\item[(ii)] replacing all empty paths through null states with effective direct transitions between non-null states,
obtaining sum-over-paths probabilities by inverting the grammar's transition matrix.
\end{description}
